### 2015-12-03 paper/1-introduction.tex

\subsection{Solution} %for lack of a better word
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the Direct connection algorithm makes reusing the adevs kernel hard.
The devs-ex-machina (dxex) kernel follows the design of the PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) kernel where appropriate, but by the very nature of the implementation languages has to differ in key aspects (e.g. memory allocation strategy). 
The core aim of the project is to offer a deterministic simulation kernel where the simulation modeller is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
\subsection{Time}
The DEVS formalisms has $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time. Dxex kernels can operate on IEEE754 floating point time units or integer time. In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages.%ref to time paper.
An integer representation significantly reduces the possible range of the virtual time, but avoids approximation errors. Furthermore, the notion of $\epsilon$ as a minimum distance between two distinct time points need not to be established. This task is non-trivial for floating point approximations.\\
The select() function, which imposes a sequential ordering between concurrent events, is made implicit in this project by extending time representation with a causality field with a range at least as large as the maximum nr of models in the kernels ($2^{24}$ by default). If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$. This avoids evaluating the select() function on all imminent models, while still maintaining the deterministic order of concurrent transitions.

>>>>

\subsection{Solution} %for lack of a better word
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the Direct connection algorithm makes reusing adevs kernel hard. Devs-ex-machina (dxex) is based on PythonPDEVS \cite{PythonPDEVS}(PyPDEVS) but differs in design and implementation.
The core aim of the project is to offer a deterministic simulation kernel where the simulation modeller is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
\subsection{Time}
The DEVS formalisms has $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time. Dxex kernels can operate on IEEE754 floating point time units or integer time. In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages.%ref to time paper.
An integer representation significantly reduces the possible range of the virtual time, but avoids approximation errors. Furthermore, the notion of $\epsilon$ as a minimum distance between two distinct time points need not to be established. This task is non-trivial for floating point approximations.\\
Our time representation is extended with a causality field with a range equal or greater than the maximum nr of models in a single kernels ($2^{24}$ by default). If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$. This allows a deterministic ordering of concurrent events. % execution not so much as tracing

### 2015-12-03 paper/2-background.tex

\subsection{Conservative Synchronization}
%TODO explain conservative synchronization, and how it uses locking to wait for continuing the simulation
Conservative synchronization is defined by the invariant that no model will advance in time before it has received all input from any influencing model. \\
This requires the concepts of earliest output time (eot) and earliest input time (eit), which define the timespan within which a model can safely advance. \\ The eit of any model is the minimum of all eot values of (directly) influencing models. A model can simulate up to (but not including) eit, then waits until that value is increased. An important disadvantage here is that the influenced-by relation is always defined at model(link) creation, not at runtime. A model that can influence another, but never does, can severely slow down the protocol.\\ 
Deadlock between models that influence each other and end up waiting on each other can be broken/avoided by a variety of means. In this simulator, the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used. \\
In our implementation null-time is the timestamp a model is guaranteed to have passed in simulation. More precisely, a null message of time t is a guarantee that any output with timestamp t-$\epsilon$ is already sent.\\
In general, the eot/eit/nulltime of a kernel is the minimum of each of those values for all models in the kernel.\\
Conservative synchronization relies explicitly on information provided by the model creator in the form of a lookahead, a relative timespan during which the model is insensitive to outside events. This lookahead can be non-trivial to calculate. In general, a simulation writer will not be able to predict the exact lookahead of models involved in an experiment without having run the experiment.\\ Conservative kernels can operate if there exists a cyclic dependency between them, but at a quite severe performance penalty, as seen in section 4.
\subsection{Optimistic Synchronization}
%TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
Optimistic synchronization allows causality errors to occur but recovers from those errors using a roll-back mechanism, the most common of which is Timewarp \cite{Jefferson:1985:VT:3916.3988}.
Whenever a kernel receives an event with a timestamp in the kernel's past, the state of the kernel (and all models) is reverted to that time. The possible gain in runtime is offset by the increase in memory required to keep saved states and (sent) messages.
In contrast to conservative, optimistic does not rely on any domain specific information. It is only sensitive to runtime use of connections, not the probability that they might occur. %reword connections ?
\\
If the (runtime) dependency graph contains a cycle, optimistic can suffer a series of cascading reverts. Without domain specific information, the kernel assumes that any event will influence at least one model, but this can lead to an infinite loop of reverts in the worst case.\\
This effect can be lessened by lazy cancellation and/or lazy re-evaluation \cite{FujimotoBook}.
\subsection{Global Virtual Time}
To avoid exhausting memory in state/event saving, optimistic synchronization relies on the concept of global virtual time\cite{Jefferson:1985:VT:3916.3988}. In optimistic simulations, GVT is defined as the lowest timestamp of any unprocessed event. \\ Intuitively this is the simulation timepoint that is certain to be preserved, corresponding exactly with the simulation up to that time in a non-parallel implementation.\\
In conservative the minimum simulation time of all kernels is the GVT, or in terms of null messages: the least timestamp of any null message in transit.
The GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, ... .

>>>>

\subsection{Conservative Synchronization}
%TODO explain conservative synchronization, and how it uses locking to wait for continuing the simulation
Conservative synchronization is defined by the invariant that no model will advance in time before it has received all input from any influencing model. \\
This requires the concepts of earliest output time (eot) and earliest input time (eit), which define the timespan within which a model can safely advance. \\ The eit of any model is the minimum of all eot values of (directly) influencing models. A model can simulate up to (but not including) eit, then waits until that value is increased. An important disadvantage here is that the influenced-by relation is always defined at model(link) creation, not at runtime. A model that can influence another, but never does, can severely slow down the protocol. 
Deadlock between models that influence each other and end up waiting on each other can be broken/avoided by a variety of means. In this simulator, the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used. \\
In our implementation null-time is the timestamp a model is guaranteed to have passed in simulation. More precisely, a null message of time t is a guarantee that any output with timestamp t-$\epsilon$ is already sent.\\
Conservative synchronization relies explicitly on information provided by the model creator in the form of a lookahead, a relative timespan during which the model is insensitive to outside events. This lookahead can be non-trivial to calculate, a simulation writer will in general not be able to determine the exact lookahead of models involved in an experiment without having run the experiment.\\ Conservative kernels can operate if there exists a cyclic dependency between them, but at a quite severe performance penalty, as seen in section 4.
\subsection{Optimistic Synchronization}
%TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
Optimistic synchronization allows causality errors to occur but recovers using a roll-back mechanism, the most common of which is Timewarp \cite{Jefferson:1985:VT:3916.3988}.
Whenever a kernel receives an event with a timestamp in the kernel's past, the state of the kernel (and all models)reverts to that time.
Because an optimistic kernel does not wait for other kernels, it can achieve a lower runtime than conservative.
The cost of this approach is higher complexity of the code and larger memory usage for state/message saving.
In contrast to conservative, optimistic does not rely on any domain specific information, it is only sensitive to runtime use of connections, not the probability that they might occur. %reword connections ?
\\
If the (runtime) dependency graph contains a cycle, optimistic can suffer a series of cascading reverts. The cost of reverts can be reduced by lazy cancellation and/or lazy re-evaluation \cite{FujimotoBook}.
\subsection{Global Virtual Time}
To avoid exhausting memory in state/event saving, optimistic synchronization relies on the concept of global virtual time\cite{Jefferson:1985:VT:3916.3988}. In optimistic simulations, GVT is defined as the lowest timestamp of any unprocessed event. \\ Intuitively this is the simulation timepoint that is certain to be preserved, corresponding exactly with a sequential simulation.
In conservative, the minimum simulation time of all kernels is the LBTS, or in terms of null messages: the least timestamp of any null message in transit.
The GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, ... .

### 2015-12-03 paper/4-performance.tex
added some comments here and there to clarify what has been done/mentioned elsewhere etc.

%TODO say that you will now prove your claims about the aproach having nearly no impact
%TODO and potentially even claim that you are faster than adevs for some situations, due to the algorithms from PythonPDEVS and further features
%TODO briefly give some hardware info on the machine (CPU, OS, RAM), used software (version of adevs, GCC, ...), and about number of samples/variance
\subsection{Sequential Simulation}
%TODO we start with sequential simulation results, because this is still base case
%TODO compare CPU time with adevs for some benchmarks
%TODO and explain why there is such a difference
%TODO (spend some time here, as this is also important)
%\subsubsection{CPU Usage}
%TODO !!! benchmarks are in integers
\subsubsection{Devstone}
The Devstone \cite{DEVStone} benchmark is highly hierarchical, using Direct connection the dxex kernels can exploit this, whereas adevs needs to walk the structure of the model to pass events.\\
\begin{figure}[h]
	\includegraphics[width=.5\textwidth]{fig/fig1.eps}
	\label{fig1.eps}
\end{figure}

@@

\subsubsection{Interconnect}
Interconnect \cite{van2013research} is a benchmark where all models broadcast, creating a complete graph in terms of dependencies between models. It highlights in sequential simulation the cost of heap allocation (messages) in dxex. As the model count increases, we see the expected quadratic increase in runtime in both kernels, but an increasing penalty for dxex w.r.t. adevs. This effect is due to the heap allocation of the messages by dxex, even though this is minimized using memory pools it remains significant.\\
\begin{figure}[h]
	\includegraphics[width=.5\textwidth]{fig/fig3.eps}
	\label{fig3.eps}
\end{figure}

@@

%TODO (keep explaining as much as space permits, as this is the evaluation of the core contribution)
\subsection{Memory Usage}
%TODO compare memory usage with adevs for some benchmarks
\subsubsection{Platform and tools}
Both dxex and adevs use tcmalloc as memory allocator. Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (malloc/free/sbrk/mmap/...). Tcmalloc will only gradually release memory back to the OS, whereas our pools will not do so at all. If memory has been allocated once, it is from a performance point of view better to keep that memory in the pool. For this reason, the memory utilization is best measured by peak allocation. Profiling is done using Valgrind's massif tool \cite{Nethercote:2007:VFH:1273442.1250746}
Platform used has an i5-3317U Intel cpu and 8GB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).\\
\subsubsection{Measure}
Adevs passes messages by value (in container by reference), we pass by pointer. The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages. This combines text, stack and heap memory for the program profiled. From the point of view of the OS or user, this is the actual memory in use. It is important to note that, especially in the case of optimistic, not all this memory is always in use by the kernel. During the simulation, the pools will generally not return memory once allocated, but keep it for later reuse.
\subsubsection{Results}
\paragraph*{Devstone}
\begin{table}[lhtb]
	\centering
	\caption{Devstone 40x40 t5e5, unit MiB, 4 kernels (if parallel)}
	\label{dtone_mem}
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		Adevs & AdevsCon &DX &DXCon &DXOpt\\ \hline
		44 & 70 & 42 & 75 & 363  \\ \hline
	\end{tabular}
\end{table}
Since conservative passes messages by pointer, it needs a GVT/LBTS implementation to organise garbage collection. This inevitable delay explains the higher memory usage compared to adevs.\\
%TODO next sentence needs rewriting
Optimistic needs a more complex GVT/garbage collection algorithm due to the Timewarp algorithm, which requires that the state of models is saved. Additionally, the differences in LP virtual times are far larger compared to conservative time synchronization. All these factors explain the heavier memory usage. Devstone (flattened) is allocated in a chain. The leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to quite severe memory pressure. Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels. 
\paragraph*{Interconnect}
In section 4.2 the parallel performance of this benchmark is further explained. It is of interest for sequential to contrast with the runtime penalty that dxex suffers w.r.t. adevs. Optimistic fails this benchmark, see section 4.2.3 for a detailed analysis.\\
\begin{table}[lhtb]
	\centering
	\caption{Interconnect w 40 t5e5, unit MiB, 2 kernels (if parallel)}
	\label{iconn_mem}
	\begin{tabular}{| l | l | l | l |}
		\hline
		Adevs & AdevsCon &DX &DXCon\\ \hline
		39 & 39 & 35 & 52  \\ \hline
	\end{tabular}
\end{table}
The discrepancy between both conservative implementations is detailed in 4.1.4.

>>>>

%TODO say that you will now prove your claims about the aproach having nearly no impact
%TODO and potentially even claim that you are faster than adevs for some situations, due to the algorithms from PythonPDEVS and further features
%TODO briefly give some hardware info on the machine (CPU, OS, RAM), used software (version of adevs, GCC, ...), and about number of samples/variance
\subsection{Sequential Simulation}
%TODO we start with sequential simulation results, because this is still base case
%TODO compare CPU time with adevs for some benchmarks
%TODO and explain why there is such a difference
%TODO (spend some time here, as this is also important)
%\subsubsection{CPU Usage}
%TODO !!! benchmarks are in integers
The benchmark models are identical in functionality (between adevs/dxex versions). Dxex uses integer timestamps, adevs can only use floating point (of the same values).
\subsubsection{Devstone}
The Devstone \cite{DEVStone} benchmark is highly hierarchical, using Direct connection the dxex kernels can exploit this, whereas adevs needs to walk the structure of the model to pass events.\\
\begin{figure}[h]
	\includegraphics[width=.5\textwidth]{fig/fig1.eps}
	\label{fig1.eps}
\end{figure}

@@

\subsubsection{Interconnect}
Interconnect \cite{van2013research} is a benchmark where all models broadcast, creating a complete graph in terms of dependencies between models. As the model count increases, we see the expected quadratic increase in runtime in both kernels, but an increasing penalty for dxex w.r.t. adevs. Profiling shows this is entirely to the heap allocation of the messages by dxex, even though minimized by using memory pools, it remains significant.\\
\begin{figure}[h]
	\includegraphics[width=.5\textwidth]{fig/fig3.eps}
	\label{fig3.eps}
\end{figure}

@@

%TODO (keep explaining as much as space permits, as this is the evaluation of the core contribution)
\subsection{Memory Usage}
%TODO compare memory usage with adevs for some benchmarks
\subsubsection{Platform and tools}
Both dxex and adevs use tcmalloc as memory allocator. Additionally, dxex uses memory pools to further reduce the frequency of expensive system calls (malloc/free/sbrk/mmap/...). Tcmalloc will only gradually release memory back to the OS, whereas our pools will not do so at all. If memory has been allocated once, it is from a performance point of view better to keep that memory in the pool. For this reason, the memory utilization is best measured by peak allocation. Profiling is done using Valgrind's massif tool \cite{Nethercote:2007:VFH:1273442.1250746}
Platform used has an i5-3317U Intel cpu and 8GB RAM with a page size of 4,096KiB, running Fedora 22 (kernel 4.2.6).\\
\subsubsection{Measure}
Adevs passes messages by value (in container by reference), we pass by pointer. The runtime effects of this choice are already demonstrated in the interconnect benchmark, so in this section, we measure memory usage in number of allocated pages. This combines text, stack and heap memory for the program profiled. From the point of view of the OS or user, this is the actual memory in use. It is important to note that, especially in the case of optimistic, not all this memory is always in use by the kernel. During the simulation, the pools will generally not return memory once allocated, but keep it for later reuse.
\subsubsection{Results}
\paragraph*{Devstone}
\begin{table}[lhtb]
	\centering
	\caption{Devstone 40x40 t5e5, unit MiB, 4 kernels (if parallel)}
	\label{dtone_mem}
	\begin{tabular}{| l | l | l | l | l |}
		\hline
		Adevs & AdevsCon &DX &DXCon &DXOpt\\ \hline
		44 & 70 & 42 & 75 & 363  \\ \hline
	\end{tabular}
\end{table}
Since conservative passes messages by pointer, it needs a GVT/LBTS implementation to organise garbage collection. This inevitable delay explains the higher memory usage compared to adevs.\\
%TODO next sentence needs rewriting --done?
Optimistic needs a more complex GVT/garbage collection algorithm due to the Timewarp algorithm, which requires that the state of models is saved. Additionally, the differences in LP virtual times are far larger compared to conservative time synchronization. All these factors explain the heavier memory usage. Devstone (flattened) is allocated in a chain. The leafs in the dependency graph will therefore do a lot of unnecessary simulation before having a revert, leading to quite severe memory pressure. Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels. 
\paragraph*{Interconnect}
In section 4.2 the parallel performance of this benchmark is further explained. It is of interest for sequential to contrast with the runtime penalty that dxex suffers w.r.t. adevs. Optimistic fails this benchmark, see section 4.2.3 for a detailed analysis.\\
\begin{table}[lhtb]
	\centering
	\caption{Interconnect w 40 t5e5, unit MiB, 2 kernels (if parallel)}
	\label{iconn_mem}
	\begin{tabular}{| l | l | l | l |}
		\hline
		Adevs & AdevsCon &DX &DXCon\\ \hline
		39 & 39 & 35 & 52  \\ \hline
	\end{tabular}
\end{table}
The discrepancy between both conservative implementations is detailed in 4.1.4.
